---
title: "Megafauna - Examination - Gray whales"
author: "Matthieu Authier, Emmanuelle Cam"
date: "2022-09-30"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1) Gray whales
Once common throughout the Northern Hemisphere, gray whales (*Eschrichtius robustus*) are now only regularly found in the North Pacific Ocean where there are two extant populations, one in the eastern and one in the western North Pacific. (<https://www.fisheries.noaa.gov/species/bowhead-whale>).

Punt and Wade (Journal of Cetacean Research and Management 12(1):15–28, 2012, <https://www.semanticscholar.org/paper/Population-status-of-the-eastern-North-Pacific-of-Punt-Wade/767d7a50bc2bc4df6c7c8729f185b0628187fe77/figure/6>) investigated the status of the eastern North Pacific stock of gray whales in 2009. We will use their data to model the population dynamics of gray whales (https://www.arlis.org/docs/vol1/B/525524065.pdf).

```{r packages}
what_u_need <- c("tidyverse", "nimble", "MCMCvis", "glue", "readxl", "matrixStats")
cran_packages <- what_u_need[!(what_u_need %in% installed.packages())]
if(length(cran_packages) != 0) {
  lapply(cran_packages, install.packages, dependencies = TRUE)
}
lapply(what_u_need, library, character.only = TRUE)

```

## Population modelling
We will model the population dynamics of gray whales assuming a Pella-Tomlinson process for density-dependence. The code _model_pt_ below is a $\texttt{nimble}$ translation of that process. First, you need to use that code as it is, and in a second time, you will make some modifications.
Do not change the function _init_pt_ for the initial values.

```{r nimble_code}
### Pella-tomlinson model: nimble code
model_pt <- nimbleCode({ 
  ## data
  # n_obs: integer, sample size
  # COUNT: integer, can include NA
  # THETA: shape of the Pella-Tomlinson process
  # prior_location_lognK: location parameter for the logN prior on K
  # prior_scale_lognK: scale parameter for the logN prior on K
  
  ## likelihood: state-space model
  # initial condition
  linpred[1] <- mu
  COUNT[1] ~ dlnorm(linpred[1] - 0.5 * sigma * sigma, sd = sigma)
  # time series
  for (t in 2:n_obs) {
    linpred[t] <- log(COUNT[t - 1] * (1 + r * (1 - pow(COUNT[t - 1] / K, THETA))))
    COUNT[t] ~ dlnorm(linpred[t] - 0.5 * sigma * sigma, sd = sigma)
  }
  
  ## priors
  # normal prior
  mu ~ dnorm(0.0, sd = 5.0)
  # half-normal prior
  sigma ~ T(dnorm(0.0, sd = 1.0), 0.0,)
  # uniform prior on maximum growth rate
  r ~ dunif(0.0, 1.0)
  # log-normal prior
  K ~ dlnorm(prior_location_lognK, sd = prior_scale_lognK)
  
  ## derived quantities: predicted abundance
  pred[1:n_obs] <- exp(linpred[1:n_obs])
})

### function to generate initial values (DO NOT MODIFY!)
init_pt <- function(chain_id, # identifier for a mcmc chain
                    y, # vector of count: the data
                    lower, # reasonable lower bound for carrying capacity K
                    upper  # reasonable upper bound for carrying capacity K
                    ) {
  COUNT <- rep(NA, length(y))
  if(any(is.na(y))) {
    COUNT[which(is.na(y))] <- rpois(sum(is.na(y)), mean(y, na.rm = TRUE))
  }
  out <- list(mu = 2.5 * rnorm(1),
              sigma = 0.5 * abs(rnorm(1)),
              r = runif(1, 0.0, 0.1),
              K = runif(1, lower, upper),
              COUNT = COUNT
              )
  return(out)
}
```

## Load the gray whale data
The data are stored in a spreadsheet called _PuntWade2010.xlsx_. Make sure your path is correct to load the data with the code below.

```{r}

gray <- readxl::read_excel(path = "PuntWade2010.xlsx",
                           sheet = 1,
                           na = "NA")

gray <- gray %>%
   mutate(abundance_cv = as.numeric(abundance_cv),
          abundance_se = sqrt(log1p(abundance_cv * abundance_cv)),
          lower = qlnorm(0.1, 
                         log(abundance_estimate) - 0.5 * abundance_se * abundance_se, 
                         abundance_se
                         ), 
          upper = qlnorm(0.9, 
                         log(abundance_estimate) - 0.5 * abundance_se * abundance_se, 
                         abundance_se
                         ),
          lower = floor(lower),
          upper = ceiling(upper)
          ) %>% print()
          
```
What are the columns _lower_ and _upper_? What is the level of confidence?

Columns upper and lower are the upper and lower bounds for the abundance estimate 
The level of confidence is 80%

## Exploratory Data Analysis
Plot the abundance data of grey whales over time using functionalities from $\texttt{ggplot}$.
```{r}

gray %>%
   ggplot(aes(x=year,y=abundance_estimate)) +
  geom_point() +
  theme_bw()
  
```
How many surveys are there (*i.e* years with data)?

```{r}

n_survey <- gray %>% 
  filter(!is.na(abundance_cv)) 

nimPrint("The number of year with data (surveys) is ", nrow(n_survey))

```

## Modelling (I)
Assume that the prior for K, the carrying capacity is a log-normal distribution of mean $20,000$ and coefficient of variation $20$%.

```{r}

 prior_scale_lognK <- log(1.20) / 2
 prior_location_lognK <- log(2E4) - 0.5 * (prior_scale_lognK)^2
 
 data.frame(K = rlnorm(1E6, prior_location_lognK, prior_scale_lognK)) %>%
   ggplot(aes(x = K)) +
   geom_histogram(breaks = seq(1E4, 3E4, 5E2)) +
   theme_bw() +
   ggtitle("Prior on K")
   
```

Fit a population dynamics model with a Pella-Tomlinson ($\theta = 2.4$) to these data using $\texttt{nimble}$ the code _model_pt_ above.

#The data constants are:
n_obs: sample size
THETA: shape of the Pella-Tomlinson process
prior_location_lognK: location parameter for the logN prior on K
prior_scale_lognK: scale parameter for the logN prior on K

#Constants are:
K : carrying capacity 
r : the maximum growth rate 
mu and sigma : initial condition for the state-space model

#The stochastic data are:
COUNT: abundance estimate

What is the likelihood of the data (*i.e* which distribution is assumed)?

Normal distribution on Count: COUNT[t] ~ dlnorm(linpred[t] - 0.5 * sigma * sigma, sd = sigma)
log normal prior on K (mean = 2E4, and CV = 20%): 
normal distribution prior on mu ~ with mean 0 and sd 5: mu ~ dnorm(0.0, sd = 5.0)
half-normal prior on sigma with mean 0 and sd 1: sigma ~ T(dnorm(0.0, sd = 1.0), 0.0,)
uniform prior on maximum growth rate: r ~ dunif(0.0, 1.0)

Make sure model parameters are all converging and that effective sample size are above $1000$ for all of them.

```{r}
## data for Nimble
data_constant <- list(n_obs = gray %>% 
                            nrow(),
                       # theta parameter of the Pella-Tomlinson process
                       THETA = 2.4,
                       # log normal prior on K (mean = 2E4, and CV = 20%)
                       prior_location_lognK =(log(2E4)-0.5*(log(1.20) / 2)^2),
                       prior_scale_lognK = log(1.20) / 2
                       )
 data_stochastic <- list(COUNT = gray %>% pull(abundance_estimate))
 
 ## specification for nimble
 n_iter <- 100000 # number of iterations
 n_burnin <- 10000 # number of burn-in
 n_chain <- 4 # number of chains
 n_thin <- 6 # thining
 n_final <- (n_iter - n_burnin) / n_thin # final sample size from posterior
```

```{r}
mcmc_pt <- nimbleMCMC(code = model_pt, 
                       constants = data_constant,
                       data = data_stochastic, 
                       inits = lapply(1:4, 
                                      init_pt, 
                                      y = gray %>%
                                        pull(abundance_estimate),
                                      lower = 1.5E4,
                                      upper = 2.5E4
                                      ),
                       nchains = n_chain, 
                       niter = n_iter,
                       nburnin = n_burnin,
                       thin = n_thin,
                       summary = FALSE, 
                       WAIC = TRUE,
                       monitors = c('mu', 'sigma', 'K', 'r', 'pred')
                       )



#Numerical summaries
 MCMCsummary(object = mcmc_pt$samples, round = 2,
             params = c('sigma', 'mu', 'K', 'r')
             )

```

```{r}
## check convergence: trace and marginal posterior density
 MCMCtrace(object = mcmc_pt$samples,
           pdf = FALSE, # no export to PDF
           ind = TRUE,  # separate density lines per chain
           params = c('sigma', 'mu', 'K', 'r')
           )
```

```{r}
## posterior analysis: extract joint posterior distribution
 posterior <- mcmc_pt$samples %>% 
   do.call("rbind", .) %>% 
   as.data.frame() %>%
   mutate(chain_id = rep(1:n_chain, each = n_final),
          iteration = rep(1:n_final, n_chain)
          )
```

How do assess that parameters' convergence has been reached?

We can look at the density lines of the chains and check if they are peaking for the same 
value, the trace must be also a combined fuzzy thread. If there are separate traces that never cross, we can assess that the parameters did not converge.
Moreover, the Rhat is closed from 0, it mean that the variance is not high beteween the different chains so parameters have converge. Additionnaly, the n.eff is high, so there is no autocorrelation. 

The following lines of code will plot the posterior distribution of K and its prior distribution.

```{r}
data.frame(posterior = posterior %>%
              pull(K)
            ) %>%
   mutate(prior = rlnorm(n(), prior_location_lognK, prior_scale_lognK)) %>%
   pivot_longer(cols = c("prior", "posterior"),
                values_to = "value",
                names_to = "param"
                ) %>%
   ggplot(aes(x = value)) +
   facet_wrap(~ param, ncol = 1) +
   geom_histogram(breaks = seq(1E4, 3E4, 5E2)) +
   theme_bw() +
   ggtitle("K")
  
```
What can you notice?

#We can see that the prior and the posterior distribution of K are not completely the same. However, the convergence has been reached for K and it's look good. So we can notice that the parameter K have converged.

What is the estimate for maximum growth rate (r)?

```{r}
posterior %>% pull(r) %>% mean()
```

Plot the posterior distribution as histogram with bins of size $0.025$.

```{r}

data.frame(posterior = posterior %>%
              pull(r)) %>%
   pivot_longer(cols = c("posterior"),
                values_to = "value",
                names_to = "param"
                ) %>%
   ggplot(aes(x = value)) +
   geom_histogram(binwidth = 0.025) +
   theme_bw() +
   ggtitle("r")

```
Is this realistic? How does that compare for example against calf production (the ratio of the number of calves over the estimated abundance)?

Compute the apparent calf production by filling in the code below:

```{r}

 gray %>%
mutate(production = calf_count / abundance_estimate) %>%
filter(!is.na(production)) %>% 
  ggplot()+
  geom_point(aes(x=year, y=production)) +
  geom_line(aes(x=year, y=production))

mean_r <- gray %>%
mutate(production = calf_count / abundance_estimate) %>%
filter(!is.na(production))
  
nimPrint("The apparent calf production is ", mean(mean_r$production))

```
It's not really realistic because the apparent calf production is very low compared to the mean posterior of r. 

With the Pella-Tomlinson($\theta = 2.4$) model, productivity is expected to be maximum at $60$% of K.
This quantity is called the MNPL (Maximum Net Productivity Level). 

What is $60$% of K? Use the posterior distribution of K.

```{r}

MNPL = posterior %>%
 mutate(MNPL = posterior %>% 
          pull(K) %>% quantile(probs = 0.6)) %>% pull(MNPL) %>% mean()

nimPrint("60% of K is ", MNPL)

```
The maximum growth rate is reached when the population is at the MNPL. Compare the apparent calf production and the inferred MNPL. 

Is the estimate of r plausible?

```{r}

posterior %>% filter(K > 20679.88 & K < 20679.89) %>% pull(r)

```
The estimate of r is not plausible because it's too high compared to the apparent calf production

## Modelling (II)

Change the prior on r to a Gamma distribution of mean $0.04$ and $95$% confidence interval between $0.01$ and $0.08$. The following $\texttt{R}$ parametrization corresponds to this prior: $\texttt{r ~ dgamma(3.645, 1/1.094E-2)}$

Simulate $1,000,000$ values from this prior with _rgamma()_ and plot them as a histogram with bins of size 0.01

```{r}
data.frame(r = rgamma(1E6, 3.645, 1/1.094E-2)) %>% 
  ggplot(aes(x = r)) +
  geom_histogram(binwidth = 0.01) + 
  theme_bw()
```

Modify the nimble code _model_pt_ to change the uniform prior on r to this gamma prior.
Call this model 'model_pt_gamr'.

```{r}

### Pella-tomlinson model: nimble code
model_pt_gamr <- nimbleCode({ 
  ## data
  # n_obs: integer, sample size
  # COUNT: integer, can include NA
  # THETA: shape of the Pella-Tomlinson process
  # prior_location_lognK: location parameter for the logN prior on K
  # prior_scale_lognK: scale parameter for the logN prior on K
  
  ## likelihood: state-space model
  # initial condition
  linpred[1] <- mu
  COUNT[1] ~ dlnorm(linpred[1] - 0.5 * sigma * sigma, sd = sigma)
  # time series
  for (t in 2:n_obs) {
    linpred[t] <- log(COUNT[t - 1] * (1 + r * (1 - pow(COUNT[t - 1] / K, THETA))))
    COUNT[t] ~ dlnorm(linpred[t] - 0.5 * sigma * sigma, sd = sigma)
  }
  
  ## priors
  # normal prior
  mu ~ dnorm(0.0, sd = 5.0)
  # half-normal prior
  sigma ~ T(dnorm(0.0, sd = 1.0), 0.0,)
  # uniform prior on maximum growth rate
  r ~ dgamma(3.645, 1/1.094E-2)
  # log-normal prior
  K ~ dlnorm(prior_location_lognK, sd = prior_scale_lognK)
  
  ## derived quantities: predicted abundance
  pred[1:n_obs] <- exp(linpred[1:n_obs])
})

### function to generate initial values (DO NOT MODIFY!)
init_pt <- function(chain_id, # identifier for a mcmc chain
                    y, # vector of count: the data
                    lower, # reasonable lower bound for carrying capacity K
                    upper  # reasonable upper bound for carrying capacity K
                    ) {
  COUNT <- rep(NA, length(y))
  if(any(is.na(y))) {
    COUNT[which(is.na(y))] <- rpois(sum(is.na(y)), mean(y, na.rm = TRUE))
  }
  out <- list(mu = 2.5 * rnorm(1),
              sigma = 0.5 * abs(rnorm(1)),
              r = runif(1, 0.0, 0.1),
              K = runif(1, lower, upper),
              COUNT = COUNT
              )
  return(out)
}

## fit the model with nimble
 mcmc_pt_gamr <- nimbleMCMC(code = model_pt_gamr, 
                       constants = data_constant,
                       data = data_stochastic, 
                       inits = lapply(1:4, 
                                      init_pt, 
                                      y = gray %>%
                                        pull(abundance_estimate),
                                      lower = 1.5E4,
                                      upper = 2.5E4
                                      ),
                       nchains = n_chain, 
                       niter = n_iter,
                       nburnin = n_burnin,
                       thin = n_thin,
                       summary = FALSE, 
                       WAIC = TRUE,
                       monitors = c('mu', 'sigma', 'K', 'r', 'pred')
                       )

## calculate numerical summaries
 MCMCsummary(object = mcmc_pt_gamr$samples, round = 2,
             params = c('sigma', 'mu', 'K', 'r')
             )
 
```

```{r}
## check convergence: trace and marginal posterior density
 MCMCtrace(object = mcmc_pt_gamr$samples,
           pdf = FALSE, # no export to PDF
           ind = TRUE,  # separate density lines per chain
           params = c('sigma', 'mu', 'K', 'r')
           )
```

Does the posterior distribution of r change? Make a histogram comparing the prior and the posterior for r. Recycle some code above to do so.

```{r}
## posterior analysis: extract joint posterior distribution
 posterior_gamr <- mcmc_pt_gamr$samples %>% 
   do.call("rbind", .) %>% 
   as.data.frame() %>%
   mutate(chain_id = rep(1:n_chain, each = n_final),
          iteration = rep(1:n_final, n_chain)
          )

 

 ##plot 
data.frame(posterior_gamr = posterior_gamr %>%
              pull(r)) %>%
   mutate(prior = rgamma(n(), 3.645, 1/1.094E-2)) %>%
   pivot_longer(cols = c("prior", "posterior_gamr"),
                values_to = "value",
                names_to = "param"
                ) %>%
   ggplot(aes(x = value)) +
   facet_wrap(~ param, ncol = 1) +
   geom_histogram(binwidth = 0.005) +
   theme_bw() +
   ggtitle("r")
   
```

We observe that the posterior distribution of r has changed and the distribution fit better with the previous one. The patterns are very closed. 

Fit this model. Ensure that all parameters have converged and that the effective sample size is large enough. Extract the joint posterior distribution.

Does your inference on K change?

No, our inference on K doesn't change

### Modelling (III)

Include in the population dynamics that this species is actually hunted.

Compute the total number of catches each year and include them in the model dynamics:

$\texttt{N[t+1] = N[t] + N[t] * r * (1 - (N[t] / K)^THETA) - CATCH[t]}$

Modify the nimble code _model_pt_gamr_ to include the catch data. Call this model _model_pt_gamr_catch_

Fit this model with nimble. Do not forget to include the catch data. Where do they go? Into the constants or into the stochastic part?

The catch data goes on the stochastic part because their are not predictible and depends of many factors. In conclusion, their are not constant so they goes in the stochastic part. 

```{r}

gray = gray %>% mutate(catch = male_removals + female_removals)

# model_pt_gamr_catch <- nimbleCode({ ... })
data_constant <- list(n_obs = gray %>% 
                            nrow(),
                       # theta parameter of the Pella-Tomlinson process
                       THETA = 2.4,
                       # log normal prior on K (mean = 2E4, and CV = 20%)
                       prior_location_lognK =(log(2E4)-0.5*(log(1.20) / 2)^2),
                       prior_scale_lognK = log(1.20) / 2
                       )
data_stochastic <- list(COUNT = gray %>% pull(abundance_estimate), 
                         CATCH = gray %>% pull(catch))
 
 
model_pt_gamr_catch <- nimbleCode({ 
  ## data
  # n_obs: integer, sample size
  # COUNT: integer, can include NA
  # THETA: shape of the Pella-Tomlinson process
  # prior_location_lognK: location parameter for the logN prior on K
  # prior_scale_lognK: scale parameter for the logN prior on K
  
  ## likelihood: state-space model
  # initial condition
  linpred[1] <- mu
  COUNT[1] ~ dlnorm(linpred[1] - 0.5 * sigma * sigma, sd = sigma)
  # time series
  for (t in 2:n_obs) {
    linpred[t] <- log(COUNT[t - 1] * (1 + r * (1 - pow(COUNT[t - 1] / K, THETA))) - CATCH[t])
    COUNT[t] ~ dlnorm(linpred[t] - 0.5 * sigma * sigma, sd = sigma)
  }
  
  ## priors
  # normal prior
  mu ~ dnorm(0.0, sd = 5.0)
  # half-normal prior
  sigma ~ T(dnorm(0.0, sd = 1.0), 0.0,)
  # uniform prior on maximum growth rate
  r ~ dgamma(3.645, 1/1.094E-2)
  # log-normal prior
  K ~ dlnorm(prior_location_lognK, sd = prior_scale_lognK)
  
  ## derived quantities: predicted abundance
  pred[1:n_obs] <- exp(linpred[1:n_obs])
})

### function to generate initial values (DO NOT MODIFY!)
init_pt <- function(chain_id, # identifier for a mcmc chain
                    y, # vector of count: the data
                    lower, # reasonable lower bound for carrying capacity K
                    upper  # reasonable upper bound for carrying capacity K
                    ) {
  COUNT <- rep(NA, length(y))
  if(any(is.na(y))) {
    COUNT[which(is.na(y))] <- rpois(sum(is.na(y)), mean(y, na.rm = TRUE))
  }
  out <- list(mu = 2.5 * rnorm(1),
              sigma = 0.5 * abs(rnorm(1)),
              r = runif(1, 0.0, 0.1),
              K = runif(1, lower, upper),
              COUNT = COUNT
              )
  return(out)
}

## specification for nimble
 n_iter <- 100000 # number of iterations
 n_burnin <- 10000 # number of burn-in
 n_chain <- 4 # number of chains
 n_thin <- 6 # thining
 n_final <- (n_iter - n_burnin) / n_thin # final sample size from posterior

## fit the model with nimble
 mcmc_pt_gamr_catch <- nimbleMCMC(code = model_pt_gamr_catch, 
                       constants = data_constant,
                       data = data_stochastic, 
                       inits = lapply(1:4, 
                                      init_pt, 
                                      y = gray %>%
                                        pull(abundance_estimate),
                                      lower = 1.5E4,
                                      upper = 2.5E4
                                      ),
                       nchains = n_chain, 
                       niter = n_iter,
                       nburnin = n_burnin,
                       thin = n_thin,
                       summary = FALSE, 
                       WAIC = TRUE,
                       monitors = c('mu', 'sigma', 'K', 'r', 'pred')
                       )
```

```{r} 
## calculate numerical summaries
 MCMCsummary(object = mcmc_pt_gamr_catch$samples, round = 2,
             params = c('sigma', 'mu', 'K', 'r')
             )

```

```{r}
## check convergence: trace and marginal posterior density
 MCMCtrace(object = mcmc_pt_gamr_catch$samples,
           pdf = FALSE, # no export to PDF
           ind = TRUE,  # separate density lines per chain
           params = c('sigma', 'mu', 'K', 'r')
           )
```
 
```{r}
 ## posterior analysis: extract joint posterior distribution
 posterior_gamr_catch <- mcmc_pt_gamr_catch$samples %>% 
   do.call("rbind", .) %>% 
   as.data.frame() %>%
   mutate(chain_id = rep(1:n_chain, each = n_final),
          iteration = rep(1:n_final, n_chain)
          )

 ##plot 
data.frame(posterior_gamr_catch = posterior_gamr_catch %>%
              pull(`pred[1]`)) %>%
   mutate(prior = rgamma(n(), 3.645, 1/1.094E-2)) %>%
   pivot_longer(cols = c("prior", "posterior_gamr_catch"),
                values_to = "value",
                names_to = "param"
                ) %>%
   ggplot(aes(x = value)) +
   facet_wrap(~ param, ncol = 1) +
   geom_histogram(binwidth = 0.005) +
   theme_bw() +
   ggtitle("r")
```

```{r}
#regarder le WAIC
mcmc_pt_gamr_catch$WAIC$WAIC
```
Ensure that all parameters have converged and that the effective sample size is large enough. Extract the joint posterior distribution.

Plot the posterior mean of the abundance according to this last model and show also the $80$% confidence interval on this plot.
Plot the raw data alongside.

```{r}

posterior_predict <- posterior_gamr_catch %>% select(.,3:44) %>%
  as.matrix()
mean_posterior <- posterior_predict %>%
  colMeans()
quantile_posterior <- posterior_predict %>%
  colQuantiles(probs = c(0.2, 0.8))
Plotzer <- cbind(gray, mean_posterior, quantile_posterior)

Plotzer %>%
  ggplot()+
  geom_linerange(aes(x = year, y = mean_posterior, ymin = `20%`, ymax = `80%`))+
  geom_point(aes(x = year, y = mean_posterior)) +
  geom_linerange(data = gray,
                aes(x = year, y = abundance_estimate, ymin = lower, ymax =
                      upper),
                color = "tomato"
                ) +
   geom_point(data = gray,
              aes(x = year, y = abundance_estimate),
              color = "tomato"
              )+
  theme_bw()+
  ggtitle("Posterior mean of the abundance and raw data ")+
  theme(legend.position="Left")+
  ylab("average abundance (number of individual)")
  
```



