---
title: "Megafauna - Examination - Gray whales"
author: "Matthieu Authier, Emmanuelle Cam"
date: "2022-09-30"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1) Gray whales
Once common throughout the Northern Hemisphere, gray whales (*Eschrichtius robustus*) are now only regularly found in the North Pacific Ocean where there are two extant populations, one in the eastern and one in the western North Pacific. (<https://www.fisheries.noaa.gov/species/bowhead-whale>).

Punt and Wade (Journal of Cetacean Research and Management 12(1):15–28, 2012, <https://www.semanticscholar.org/paper/Population-status-of-the-eastern-North-Pacific-of-Punt-Wade/767d7a50bc2bc4df6c7c8729f185b0628187fe77/figure/6>) investigated the status of the eastern North Pacific stock of gray whales in 2009. We will use their data to model the population dynamics of gray whales (https://www.arlis.org/docs/vol1/B/525524065.pdf).

```{r packages}
what_u_need <- c("tidyverse", "nimble", "MCMCvis", "glue")
cran_packages <- what_u_need[!(what_u_need %in% installed.packages())]
if(length(cran_packages) != 0) {
  lapply(cran_packages, install.packages, dependencies = TRUE)
}
lapply(what_u_need, library, character.only = TRUE)

```

Warnings: to get the full grade for this examn, use the tidy grammar. Non-tidy solutions will not earn a full grade.

## Population modelling
We will model the population dynamics of gray whales assuming a Pella-Tomlinson process for density-dependence. The code _model_pt_ below is a $\texttt{nimble}$ translation of that process. First, you need to use that code as it is, and in a second time, you will make some modifications.
Do not change the function _init_pt_ for the initial values.

```{r nimble_code}
### Pella-tomlinson model: nimble code
model_pt <- nimbleCode({ 
  ## data
  # n_obs: integer, sample size
  # COUNT: integer, can include NA
  # THETA: shape of the Pella-Tomlinson process
  # prior_location_lognK: location parameter for the logN prior on K
  # prior_scale_lognK: scale parameter for the logN prior on K
  
  ## likelihood: state-space model
  # initial condition
  linpred[1] <- mu
  COUNT[1] ~ dlnorm(linpred[1] - 0.5 * sigma * sigma, sd = sigma)
  # time series
  for (t in 2:n_obs) {
    linpred[t] <- log(COUNT[t - 1] * (1 + r * (1 - pow(COUNT[t - 1] / K, THETA))))
    COUNT[t] ~ dlnorm(linpred[t] - 0.5 * sigma * sigma, sd = sigma)
  }
  
  ## priors
  # normal prior
  mu ~ dnorm(0.0, sd = 5.0)
  # half-normal prior
  sigma ~ T(dnorm(0.0, sd = 1.0), 0.0,)
  # uniform prior on maximum growth rate
  r ~ dunif(0.0, 1.0)
  # log-normal prior
  K ~ dlnorm(prior_location_lognK, sd = prior_scale_lognK)
  
  ## derived quantities: predicted abundance
  pred[1:n_obs] <- exp(linpred[1:n_obs])
})

### function to generate initial values (DO NOT MODIFY!)
init_pt <- function(chain_id, # identifier for a mcmc chain
                    y, # vector of count: the data
                    lower, # reasonable lower bound for carrying capacity K
                    upper  # reasonable upper bound for carrying capacity K
                    ) {
  COUNT <- rep(NA, length(y))
  if(any(is.na(y))) {
    COUNT[which(is.na(y))] <- rpois(sum(is.na(y)), mean(y, na.rm = TRUE))
  }
  out <- list(mu = 2.5 * rnorm(1),
              sigma = 0.5 * abs(rnorm(1)),
              r = runif(1, 0.0, 0.1),
              K = runif(1, lower, upper),
              COUNT = COUNT
              )
  return(out)
}
```

## Load the gray whale data
The data are stored in a spreadsheet called _PuntWade2010.xlsx_. Make sure your path is correct to load the data with the code below.

```{r}

library(readxl)
library(nimble)

gray <- PuntWade2010

gray %>%
   mutate(abundance_cv = as.numeric(abundance_cv),
          abundance_se = sqrt(log1p(abundance_cv * abundance_cv)),
          lower = qlnorm(0.1, 
                         log(abundance_estimate) - 0.5 * abundance_se * abundance_se, 
                         abundance_se
                         ), 
          upper = qlnorm(0.9, 
                         log(abundance_estimate) - 0.5 * abundance_se * abundance_se, 
                         abundance_se
                         ),
          lower = floor(lower),
          upper = ceiling(upper)
          )
```
?qlnorm
What are the columns _lower_ and _upper_? What is the level of confidence?

Upper and lower bounds for carrying capacity
20% 

## Exploratory Data Analysis
Plot the abundance data of grey whales over time using functionalities from $\texttt{ggplot}$.
#Y'a untruc plus stylé à faire là je pense
```{r}
gray %>%
   ggplot(aes(x=year,y=abundance_estimate)) +
  geom_point()+
  theme_bw()
```

How many surveys are there (*i.e* years with data)?

```{r}
# write your answer in R here
n_survey <- gray %>% 
  filter(!is.na(abundance_cv)) 

nimPrint("Answear is ", nrow(n_survey))


```

## Modelling (I)
Assume that the prior for K, the carrying capacity is a log-normal distribution of mean $20,000$ and coefficient of variation $20$%.

```{r}
 prior_scale_lognK <- log(1.20) / 2
 prior_location_lognK <- log(2E4) - 0.5 * (prior_scale_lognK)^2
 
 data.frame(K = rlnorm(1E6, prior_location_lognK, prior_scale_lognK)) %>%
   ggplot(aes(x = K)) +
   geom_histogram(breaks = seq(1E4, 3E4, 5E2)) +
   theme_bw() +
   ggtitle("Prior on K")
```

Fit a population dynamics model with a Pella-Tomlinson ($\theta = 2.4$) to these data using $\texttt{nimble}$ the code _model_pt_ above.

What are the constants?

-> parameters are 'mu', 'sigma', 'K', 'r'
( à détailler)

What are the stochastic data?

list(COUNT = gray %>% pull(abundance_estimate))

What is the likelihood of the data (*i.e* which distribution is assumed)?

log normal prior on K (mean = 2E4, and CV = 20%)

Make sure model parameters are all converging and that effective sample size are above $1000$ for all of them.

```{r}
## data for Nimble
data_constant <- list(n_obs = gray %>% 
                            nrow(),
                       # theta parameter of the Pella-Tomlinson process
                       THETA = 2.4,
                       # log normal prior on K (mean = 2E4, and CV = 20%)
                       prior_location_lognK =(log(2E4)-0.5*(log(1.20) / 2)^2),
                       prior_scale_lognK = log(1.20) / 2
                       )
 data_stochastic <- list(COUNT = gray %>% pull(abundance_estimate))
 
 ## specification for nimble
 n_iter <- 100000 # number of iterations
 n_burnin <- 10000 # number of burn-in
 n_chain <- 4 # number of chains
 n_thin <- 6 # thining
 n_final <- (n_iter - n_burnin) / n_thin # final sample size from posterior
```

Once you have completed the lines of code above, run the lines of code below (remove the #)

```{r}
## fit the model with nimble
 mcmc_pt <- nimbleMCMC(code = model_pt, 
                       constants = data_constant,
                       data = data_stochastic, 
                       inits = lapply(1:4, 
                                      init_pt, 
                                      y = gray %>%
                                        pull(abundance_estimate),
                                      lower = 1.5E4,
                                      upper = 2.5E4
                                      ),
                       nchains = n_chain, 
                       niter = n_iter,
                       nburnin = n_burnin,
                       thin = n_thin,
                       summary = FALSE, 
                       WAIC = TRUE,
                       monitors = c('mu', 'sigma', 'K', 'r', 'pred')
                       )

## calculate numerical summaries
 MCMCsummary(object = mcmc_pt$samples, round = 2,
             params = c('sigma', 'mu', 'K', 'r')
             )

## check convergence: trace and marginal posterior density
 MCMCtrace(object = mcmc_pt$samples,
           pdf = FALSE, # no export to PDF
           ind = TRUE,  # separate density lines per chain
           params = c('sigma', 'mu', 'K', 'r')
           )

## posterior analysis: extract joint posterior distribution
 posterior <- mcmc_pt$samples %>% 
   do.call("rbind", .) %>% 
   as.data.frame() %>%
   mutate(chain_id = rep(1:n_chain, each = n_final),
          iteration = rep(1:n_final, n_chain)
          )
```

How do assess that parameters' convergence has been reached?

The following lines of code will plot the posterior distribution of K and its prior distribution.

What can you notice?

```{r}
data.frame(posterior = posterior %>%
              pull(K)
            ) %>%
   mutate(prior = rlnorm(n(), prior_location_lognK, prior_scale_lognK)) %>%
   pivot_longer(cols = c("prior", "posterior"),
                values_to = "value",
                names_to = "param"
                ) %>%
   ggplot(aes(x = value)) +
   facet_wrap(~ param, ncol = 1) +
   geom_histogram(breaks = seq(1E4, 3E4, 5E2)) +
   theme_bw() +
   ggtitle("K")
  
```

What is the estimate for maximum growth rate (r)?
Plot the posterior distribution as histogram with bins of size $0.025$.

```{r}

data.frame(posterior = posterior %>%
              pull(r)) %>%
   pivot_longer(cols = c("posterior"),
                values_to = "value",
                names_to = "param"
                ) %>%
   ggplot(aes(x = value)) +
   geom_histogram(binwidth = 0.025) +
   theme_bw() +
   ggtitle("r")

```

Is this realistic? How does that compare for example against calf production (the ratio of the number of calves over the estimated abundance)?
Compute the apparent calf production by filling in the code below:

```{r}
 gray %>%
mutate(production = calf_count / abundance_estimate) %>%
filter(!is.na(production)) %>% 
  ggplot()+
  geom_point(aes(x=year, y=production)) +
  geom_line(aes(x=year, y=production))

mean_r <- gray %>%
mutate(production = calf_count / abundance_estimate) %>%
filter(!is.na(production))
  
nimPrint("Answer is ", mean(mean_r$production))

```

With the Pella-Tomlinson($\theta = 2.4$) model, productivity is expected to be maximum at $60$% of K.
This quantity is called the MNPL (Maximum Net Productivity Level). 

What is $60$% of K? Use the posterior distribution of K.

```{r}
MNPL = posterior %>%
 mutate(MNPL = posterior %>% 
          pull(K) %>% quantile(probs = 0.6)) %>% pull(MNPL) %>% mean()

nimPrint("60% of K is ", MNPL)
```

The maximum growth rate is reached when the population is at the MNPL. Compare the apparent calf production and the inferred MNPL. 
Is the estimate of r plausible?

## Modelling (II)
Change the prior on r to a Gamma distribution of mean $0.04$ and $95$% confidence interval between $0.01$ and $0.08$. The following $\texttt{R}$ parametrization corresponds to this prior: $\texttt{r ~ dgamma(3.645, 1/1.094E-2)}$

Simulate $1,000,000$ values from this prior with _rgamma()_ and plot them as a histogram with bins of size 0.01

```{r}
# data.frame(r = rgamma(1E6, 3.645, 1/1.094E-2)) %>%
#   ... +
#   theme_bw()
```

Modify the nimble code _model_pt_ to change the uniform prior on r to this gamma prior.
Call this model 'model_pt_gamr'.

```{r}
# model_pt_gamr <- nimbleCode({ ... })
```

Fit this model. Ensure that all parameters have converged and that the effective sample size is large enough. Extract the joint posterior distribution.

Does the posterior distribution of r change? Make a histogram comparing the prior and the posterior for r. Recycle some code above to do so.

Does your inference on K change?

### Modelling (III)
Include in the population dynamics that this species is actually hunted.

Compute the total number of catches each year and include them in the model dynamics:

$\texttt{N[t+1] = N[t] + N[t] * r * (1 - (N[t] / K)^THETA) - CATCH[t]}$

Modify the nimble code _model_pt_gamr_ to include the catch data. Call this model _model_pt_gamr_catch_

```{r}
# model_pt_gamr_catch <- nimbleCode({ ... })
```

Fit this model with nimble. Do not forget to include the catch data. Where do they go? Into the constants or into the stochastic part?

Ensure that all parameters have converged and that the effective sample size is large enough. Extract the joint posterior distribution.

Plot the posterior mean of the abundance according to this last model and show also the $80$% confidence interval on this plot.
Plot the raw data alongside.
Hint: use the following code:

```{r}
# geom_linerange(data = gray,
#                aes(x = year, y = abundance_estimate, ymin = lower, ymax = upper),
#                color = "tomato"
#                ) +
#   geom_point(data = gray,
#              aes(x = year, y = abundance_estimate),
#              color = "tomato"
#              )
```



